Appendix A

### 2023-11-13 manuscript/manuscript.Rmd
New mle-based optimisation passes checks and yields promising output.
Manuscript revision.

require(power.transform)
require(ggplot2)
require(patchwork)
require(data.table)
require(latex2exp)
require(coro)

@@

-   We devise shift-sensitive versions of the Box-Cox and Yeo-Johnson transformation, including versions robust to outliers.

-   We define an empirical goodness of fit test for detecting cases where power transformations fail to yield a normally distributed transformed feature.

@@

## Shift-sensitive power transformation

Box-Cox and Yeo-Johnson transformations are modified by introducing a shift parameter $x_0$ into equations \ref{eqn:box-cox-original} and \ref{eqn:yeo-johnson-original}. The shift-sensitive Box-Cox transformation of a feature value $x_i$ of feature $\mathbf{X}$ under transformation parameter $\lambda$ and shift parameter $x_0$ is then:

```{=latex}
\begin{equation}
\label{eqn:box-cox-shift}
\phi_{\text{BC}}^{\lambda, x_0} (x_i) = 
\begin{cases}
\left( \left(x_i - x_0 \right)^\lambda - 1 \right) / \lambda & \text{if } \lambda \neq 0\\
\log(x_i - x_0) & \text{if } \lambda = 0
\end{cases}
\end{equation}
```
Where $x_i - x_0 > 0$. Likewise, the shift-sensitive Yeo-Johnson transformation of a feature value $x_i$ under transformation parameter $\lambda$ and shift parameter $x_0$ is:

```{=latex}
\begin{equation}
\label{eqn:yeo-johnson-shift}
\phi_{\text{YJ}}^{\lambda, x_0} (x_i) = 
\begin{cases}
\left( \left( 1 + x_i - x_0\right)^\lambda - 1\right) / \lambda & \text{if } \lambda \neq 0 \text{ and } x_i - x_0 \geq 0\\
\log(1 + x_i - x_0) & \text{if } \lambda = 0 \text{ and } x_i - x_0 \geq 0\\
-\left( \left( 1 - x_i - x_0\right)^{2 - \lambda} - 1 \right) / \left(2 - \lambda \right) & \text{if } \lambda \neq 2 \text{ and } x_i - x_0< 0\\
-\log(1 - x_i - x_0) & \text{if } \lambda = 2 \text{ and } x_i - x_0 < 0
\end{cases}
\end{equation}
```
For both shift-sensitive transformations, $\lambda$ and $x_0$ are obtained by maximising the normal log-likelihood function. For the shift-sensitive Box-Cox transformation, and ignoring constant terms, this can be written as:

```{=latex}
\begin{equation}
\label{eqn:box-cox-shift-log-likelihood}
\left\{ \hat{\lambda}^{\text{BC}}, \hat{x}_0^{\text{BC}} \right\} = \argmax_{\lambda, x_0} \sum_{i=1}^n -\frac{1}{2}\log(\sigma^2) + (\lambda - 1) \log(x_i - x_0)
\end{equation}
```
where $\sigma^2$ is the variance of the Box-Cox transformed feature $\phi_{\text{BC}}^{\lambda, x_0} (\mathbf{X})$. Similarly for the shift-sensitive Yeo-Johnson transformation:

```{=latex}
\begin{equation}
\label{eqn:yeo-johnson-shift-log-likelihood}
\left\{ \hat{\lambda}^{\text{YJ}}, \hat{x}_0^{\text{YJ}} \right\} = \argmax_{\lambda, x_0} \sum_{i=1}^n -\frac{1}{2}\log(\sigma^2) + (\lambda - 1) \sgn(x_i - x_0) \log(1 + |x_i - x_0|)
\end{equation}
```
where $\sigma^2$ is the variance of Yeo-Johnson transformed feature $\phi_{\text{YJ}}^{\lambda, x_0} (\mathbf{X})$.

>>>>

require(power.transform)
require(ggplot2)
require(patchwork)
require(data.table)
require(latex2exp)
require(coro)
require(paletteer)
require(robustHD)
require(modeldata)

@@

-   We devise shift and scale invariant versions of the Box-Cox and Yeo-Johnson transformation, including versions robust to outliers.

-   We define an empirical central normality test for detecting cases where power transformations fail to yield a roughly normally distributed transformed feature.

@@

## Shift and scale invariant power transformation

Box-Cox and Yeo-Johnson transformations are modified by introducing shift parameter $x_0$ and scale parameter $s$ into equations \ref{eqn:box-cox-original} and \ref{eqn:yeo-johnson-original}. The shift and scale invariant Box-Cox transformation of a feature value $x_i$ of feature $\mathbf{X}$ under transformation parameter $\lambda$, shift parameter $x_0$ and scale parameter $s$ is then:

```{=latex}
\begin{equation}
\label{eqn:box-cox-shift}
\phi_{\text{BC}}^{\lambda, x_0, s} (x_i) = 
\begin{cases}
\left( \left(\frac{x_i - x_0}{s} \right)^\lambda - 1 \right) / \lambda & \text{if } \lambda \neq 0\\
\log\left[\frac{x_i - x_0}{s}\right] & \text{if } \lambda = 0
\end{cases}
\end{equation}
```

Where $x_i - x_0 > 0$. Likewise, the shift and scale invariant Yeo-Johnson transformation of a feature value $x_i$ under transformation parameter $\lambda$, shift parameter $x_0$ and scale parameter $s$ is:

```{=latex}
\begin{equation}
\label{eqn:yeo-johnson-shift}
\phi_{\text{YJ}}^{\lambda, x_0, s} (x_i) = 
\begin{cases}
\left( \left( 1 + \frac{x_i - x_0}{s}\right)^\lambda - 1\right) / \lambda & \text{if } \lambda \neq 0 \text{ and } x_i - x_0 \geq 0\\
\log\left[1 + \frac{x_i - x_0}{s}\right] & \text{if } \lambda = 0 \text{ and } x_i - x_0 \geq 0\\
-\left( \left( 1 - \frac{x_i - x_0}{s}\right)^{2 - \lambda} - 1 \right) / \left(2 - \lambda \right) & \text{if } \lambda \neq 2 \text{ and } x_i - x_0 < 0\\
-\log\left[1 - \frac{x_i - x_0}{s}\right] & \text{if } \lambda = 2 \text{ and } x_i - x_0 < 0
\end{cases}
\end{equation}
```

For both invariant transformations, $\lambda$, $x_0$ and $s$ are obtained by maximising the log-likelihood function. A full derivation is shown in Appendix A. The invariant Box-Cox log-likelihood function is:

```{=latex}
\begin{equation}
\begin{split}
\mathcal{l}_{\text{BC}}^{\lambda, x_0, s} = & -\frac{n}{2} \log \left[2 \pi \sigma^2 \right] -\frac{1}{2 \sigma^2} \sum_{i=1}^n \left( \phi^{\lambda, x_0, s}(x_i) - \mu \right)^2 \\
& -n \lambda \log s + \left( \lambda - 1 \right) \sum_{i=1}^n \log \left[ x_i - x_0 \right]
\end{split}
\end{equation}
```
where $\mu$ and $\sigma^2$ are the mean and variance of the Box-Cox transformed feature $\phi_{\text{BC}}^{\lambda, x_0} (\mathbf{X})$, respectively. Similarly, for the invariant Yeo-Johnson log-likelihood function is:
HEREHEHEHHRHEHHEHEHRER
```{=latex}
\begin{equation}
\label{eqn:yeo-johnson-shift-log-likelihood}
\left\{ \hat{\lambda}^{\text{YJ}}, \hat{x}_0^{\text{YJ}} \right\} = \argmax_{\lambda, x_0} \sum_{i=1}^n -\frac{1}{2}\log(\sigma^2) + (\lambda - 1) \sgn(x_i - x_0) \log(1 + |x_i - x_0|)
\end{equation}
```
where $\sigma^2$ is the variance of Yeo-Johnson transformed feature $\phi_{\text{YJ}}^{\lambda, x_0} (\mathbf{X})$.

### 2023-11-13 manuscript/manuscript_appendix_a.Rmd
Update manuscript_appendix_a.Rmd
Updated log likelihood to correct formulation.
Manuscript revision.

@@

This leads to the following log-likelihood for the Box-Cox transformation:
```{=latex}
\begin{equation}
\begin{split}
\mathcal{l}_{\text{BC}}^{\lambda, x_0, s} = & -\frac{n}{2} \log{2 \pi \sigma^2} -\frac{1}{2 \sigma^2} \sum_{i=1}^n \left( \phi^{\lambda, x_0, s}(x_i) - \mu \right)^2 \\
& -n \lambda \log s + \left( \lambda - 1 \right) \sum_{i=1}^n \log \left[ x_i - x_0 \right]
\end{split}
\end{equation}
```

@@

This lead to the following log-likelihood for the Yeo-Johnson transformation:
```{=latex}
\begin{equation}
\begin{split}
\mathcal{l}_{\text{BC}}^{\lambda, x_0, s} = & -\frac{n}{2} \log{2 \pi \sigma^2} -\frac{1}{2 \sigma^2} \sum_{i=1}^n \left( \phi^{\lambda, x_0, s}(x_i) - \mu \right)^2 \\
& - n \log s + (\lambda - 1) \sum_{i=1}^n \sgn(x_i - x_0) \log \left[1 + \frac{|x_i - x_0|}{s} \right]
\end{split}
\end{equation}
```

>>>>

## Box-Cox transformation

@@

This leads to the following log-likelihood for the Box-Cox transformation:
```{=latex}
\begin{equation}
\begin{split}
\mathcal{l}_{\text{BC}}^{\lambda, x_0, s} = & -\frac{n}{2} \log \left[2 \pi \sigma^2 \right] -\frac{1}{2 \sigma^2} \sum_{i=1}^n \left( \phi^{\lambda, x_0, s}(x_i) - \mu \right)^2 \\
& -n \lambda \log s + \left( \lambda - 1 \right) \sum_{i=1}^n \log \left[ x_i - x_0 \right]
\end{split}
\end{equation}
```

We introduced per-sample weights $w_i$ to facilitate robust power transformations. The log likelihood of the robust Box-Cox transformation is:
```{=latex}
\begin{equation}
\begin{split}
\mathcal{l}_{\text{rBC}}^{\lambda, x_0, s} = & -\frac{1}{2} \left(\sum_{i=1}^n w_i \right) \log \left[ 2 \pi \sigma_w^2 \right] -\frac{1}{2 \sigma_w^2} \sum_{i=1}^n w_i \left( \phi^{\lambda, x_0, s}(x_i) - \mu_w \right)^2 \\
& - \lambda \left( \sum_{i=1}^n w_i \right) \log s + \left( \lambda - 1 \right) \sum_{i=1}^n w_i \log \left[ x_i - x_0 \right]
\end{split}
\end{equation}
```
where $\sigma^2_w$ is the weighted variance of the Box-Cox transformed feature $\phi_{\text{BC}}^{\lambda, x_0} (\mathbf{X})$:

```{=latex}
\begin{equation}
\sigma_w^2 = \frac{\sum_{i=1}^n w_i \left(\phi_{\text{BC}}^{\lambda, x_0, s} (x_i) - \mu_w \right)^2}{\sum_{i=1}^n w_i} \quad \text{with } \mu_w = \frac{\sum_{i=1}^n \phi_{\text{BC}}^{\lambda, x_0, s} (x_i)} {\sum_{i=1}^n w_i}
\end{equation}
```

## Yeo-Johnson transformation

@@

This lead to the following log-likelihood for the Yeo-Johnson transformation:
```{=latex}
\begin{equation}
\begin{split}
\mathcal{l}_{\text{YJ}}^{\lambda, x_0, s} = & -\frac{n}{2} \log\left[2 \pi \sigma^2\right] -\frac{1}{2 \sigma^2} \sum_{i=1}^n \left( \phi^{\lambda, x_0, s}(x_i) - \mu \right)^2 \\
& - n \log s + (\lambda - 1) \sum_{i=1}^n \sgn(x_i - x_0) \log \left[1 + \frac{|x_i - x_0|}{s} \right]
\end{split}
\end{equation}
```

The log likelihood of the robust Yeo-Johnson transformation is:
```{=latex}
\begin{equation}
\begin{split}
\mathcal{l}_{\text{rYJ}}^{\lambda, x_0, s} = & -\frac{1}{2} \left(\sum_{i=1}^n w_i \right) \log \left[ 2 \pi \sigma_w^2 \right] -\frac{1}{2 \sigma_w^2} \sum_{i=1}^n w_i \left( \phi^{\lambda, x_0, s}(x_i) - \mu_w \right)^2 \\
& - \left( \sum_{i=1}^n w_i \right) \log s + (\lambda - 1) \sum_{i=1}^n w_i \sgn(x_i - x_0) \log \left[1 + \frac{|x_i - x_0|}{s} \right]
\end{split}
\end{equation}
```
where $\sigma^2_w$ is the weighted variance of the Yeo-Johnson transformed feature $\phi_{\text{BC}}^{\lambda, x_0} (\mathbf{X})$:

```{=latex}
\begin{equation}
\sigma_w^2 = \frac{\sum_{i=1}^n w_i \left(\phi_{\text{YJ}}^{\lambda, x_0, s} (x_i) - \mu_w \right)^2}{\sum_{i=1}^n w_i} \quad \text{with } \mu_w = \frac{\sum_{i=1}^n \phi_{\text{YJ}}^{\lambda, x_0, s} (x_i)} {\sum_{i=1}^n w_i}
\end{equation}
```
